Step by Step: Using Step Selection Analysis to Simulate Dispersal and Assess Landscape Connectivity

### 2021-02-25 Chapter_1/04_Manuscript/Manuscript.tex
Worked on revision
back to old data
Various updates

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Step by Step: Using Integrated Step Selection Analysis to Simulate Wild
Dog Dispersal and Assess Landscape Connectivity}

@@

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on free-ranging wild
dogs inhabiting the Okavango Delta in northern Botswana. We identified potential
dispersers based on criteria reported in \cite{Behr.2020}, immobilized them
according to protocols described in \cite{Osofsky.1996}, and fitted them with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}). Handling and collaring of all individuals was carried out and
supervised by a Botswana-registered wildlife veterinarian. Of all collared
individuals, a total of 16 individuals dispersed, each in a same-sex coalition
(7 female and 9 male coalitions). When we observed dispersal we remotely
increased the GPS fixrate from 24 to 4 hours. Collected relocations were then
regularly transmitted via Iridium satellite system to a base station. This
allowed remote tracking of dispersers even if they left the main study area or
crossed international borders. To effectively distinguish between resident and
dispersersing individuals, we applied the net-squared displacement metric. This
metric measures the Euclidean distance of a collared individual to a reference
point \citep{Borger.2012}. In our case, these reference points resembled the
center of the dispersing coalition's natal home range. Hence, dispersal was
deemed to have started when an individual left its natal home range and ended
when the individual became stationary again. Previous research found no
differences between males and females during dispersal \citep{Woodroffe.2019,
Cozzi.2020}, so we did not distinguish sexes in our analyses. GPS relocations
were then converted to steps, where a step represented the straight-line
distance travelled between two consequtive GPS relocations
\citep{Turching.1998}.

@@

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In the iSSF framework, observed steps
receive a selection score \(w(x)\) that depends on the covariates \(X\)
experienced along a step, as well as the animals preferences towards these
covariates \(\beta\). In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be modelled. Thus, the
method produces more accurate selection estimates and allows to use the
resulting models as proper mechanistic movement models from which movement can
be generated. To conduct iSSf analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
changes in inference when sampling further random steps. Along each step we
extracted spatial covariates using the \textit{velox} package. We also
calculated step metrics, namely the natural logarithm of the step length
\(log(sl\_\)) and the cosine of the turning angle \(cos(ta\_)\). We scaled all
continuous covariates to a mean of zero and a standard deviation of one. We then
used the r-package \textit{glmmTMB} to fit mixed effects conditional logistic
regression models as proposed by \citep{Muff.2020}. Our movement model was built
around a previously published habitat selection model of dispersing wild dogs.
Because the earlier model was used to predict landscape permeability,
interactions between environmental and spatial covariates were impossible to
model. Our current movement model, in contrast, allowed the inclusion of
additional interactions that describe differences in movement behavior induced
through different environments. Hence, we started with the habitat model and
iteratively increased model complexity by including additional interactions
between movement metrics and environmental covariates. That is, we proposed all
two-way interactions between spatial covariates and movement metrics. For
instance, for the covariate \textit{water} we proposed the interactions
Water:cos(ta\_) and Water:log(sl\_). Despite interactions between environmental
factors and movement metrics, we also proposed the interaction
\(log(sl\_):MainActivity\) to account for the diurnal activity pattern of wild
dogs. A preliminary revealed little differences in turning angles during main
and low activity, hence we did not include the interaction
\(cos(ta\_):MainActivity\). We then ran stepwise modal forward selection absed
on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}) values and
identified the most parsimonious movement model.

>>>>

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Step by Step: Using Step Selection Analysis to Simulate Dispersal and
Assess Landscape Connectivity}

@@

\subsection{GPS Relocation Data}
Between 2011 and 2020, we collected GPS relocation data on free-ranging wild
dogs inhabiting the Okavango Delta in northern Botswana. We identified potential
dispersers based on criteria reported in \cite{Behr.2020}, immobilized them
according to protocols described in \cite{Osofsky.1996}, and fitted them with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}). Handling and collaring of all individuals was carried out and
supervised by a Botswana-registered wildlife veterinarian. Of all collared
individuals, a total of 16 individuals dispersed, each in a same-sex coalition
(7 female and 9 male coalitions). When we observed dispersal we remotely
increased the GPS fixrate from 24 to 4 hours. Collected relocations were then
regularly transmitted via Iridium satellite system to a base station. This
allowed remote tracking of dispersers even if they left the main study area or
crossed international borders. To effectively distinguish between resident and
dispersersing individuals, we applied the net-squared displacement metric. This
metric measures the Euclidean distance of a collared individual to a reference
point \citep{Borger.2012}. In our case, these reference points resembled the
center of the dispersing coalition's natal home range. Hence, dispersal was
deemed to have started when an individual left its natal home range and ended
when the individual became stationary again. Previous research found no
differences between males and females during dispersal \citep{Woodroffe.2019,
Cozzi.2020}, so we did not distinguish sexes in our analyses. GPS relocations
were then converted to steps, where a step represented the straight-line
distance travelled between two consequtive GPS relocations
\citep{Turchin.1998}.

@@

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In the iSSF framework, observed steps
receive a selection score \(w(x)\) that depends on the covariates \(X\)
experienced along a step, as well as the animals preferences towards these
covariates \(\beta\). In contrast to regular step selection analysis,
\textit{integrated} step selection analysis allows simulatneous inference on
movement and habitat preferences of the studied animal. Moreover, potential
interactions between habitat and movement preferences can be modelled. Thus, the
method produces more accurate selection estimates and allows to use the
resulting models as proper mechanistic movement models from which movement can
be generated. To conduct iSSf analysis, we paired each observed step with 24
random steps. Random steps basically resembled potential alternatives that the
animal could have realized but decided not to. We generated random steps by
sampling random turning angles from a uniform distribution (\(-\pi, +\pi\)) and
step lengths from a gamma distirbution that was fitted to observed steps (scale
= 6308, shape = 0.37). While the number of random steps is inversely
proportional to the sampling error \citep{Avgar.2016}, we found only minor
changes in inference when sampling further random steps. Along each step we
extracted spatial covariates using the \textit{velox} package. We also
calculated step metrics, namely the natural logarithm of the step length
\(log(sl\_\)) and the cosine of the turning angle \(cos(ta\_)\). We scaled all
continuous covariates to a mean of zero and a standard deviation of one. We then
used the r-package \textit{glmmTMB} to fit mixed effects conditional logistic
regression models as proposed by \citep{Muff.2020}. Our movement model was built
around a previously published habitat selection model of dispersing wild dogs.
Because the earlier model was used to predict landscape permeability,
interactions between environmental and spatial covariates were impossible to
model. Our current movement model, in contrast, allowed the inclusion of
additional interactions that describe differences in movement behavior induced
through different environments. Hence, we started with the habitat model and
iteratively increased model complexity by including additional interactions
between movement metrics and environmental covariates. That is, we proposed all
two-way interactions between spatial covariates and movement metrics. For
instance, for the covariate \textit{water} we proposed the interactions
Water:cos(ta\_) and Water:log(sl\_). Despite interactions between environmental
factors and movement metrics, we also proposed the interaction
\(log(sl\_):MainActivity\) to account for the diurnal activity pattern of wild
dogs. A preliminary revealed little differences in turning angles during main
and low activity, hence we did not include the interaction
\(cos(ta\_):MainActivity\). We then ran stepwise model forward selection based
on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}) values and
identified the most parsimonious movement model. Although several models
received an AIC weight above one, we decided to only consider the  ``best''
model for simplicity. In either way, the models with positive weights contained
almost identical covariates and an averaged model would have only given weak
support to additional covariates.
